### 2019-08-25 manuscript.Rmd
manuscript

Sensitivity analysis (SA) is a mathematical technique to investigate how variations in model parameters affect model outputs [@saltelli2008global]. An increasing number of studies use SA approach to determine which model parameters contribute to high variation in model predictions [@ferretti2016trends]. This technique has also been applied in pharmacology and toxicology research [@loizou2015application; @mcnally2012reconstruction]. Pharmacokinetic models describe the changes in the concentrations or amounts of a substance in the body over time (the various terms for the kinetic models, including "pharmacokinetic", "toxicokinetic", and "biokinetic" models, are used exchangeably here). The goal of SA in pharmacokinetic research is to examine the sensitivity of output variables,such as chemical concentration in blood or tissues, with respect to input parameters, such as anatomical, physiological, and kinetic constants [@mcnally2011workflow]. In addition, this SA approach can guide experimental design and the parameter estimation processes [@zhang2015sobol]. For instance, SA can identify "unidentifiable" parameters that can lead to problems with numerical procedures used in parameter estimation [@chu2010quantitative]. It can be further applied to parameter prioritization and parameter fixing before model calibration [@fphar201800588].

In our earlier work [@fphar201800588], we developed an updated workflow to apply global SA to reduce the computational burden in the Bayesian, Markov Chain Monte Carlo (MCMC)-based calibration process of a physiologically based pharmacokinetic (PBPK) model. We used \pkg{GNU MCSim} [@JSSv002i09], an effective simulation package for Bayesian population PBPK modeling, to calibrate the model. We found that the extended Fourier Amplitude Sensitivity Test (eFAST), a type of variance-based global SA algorithm, had the best balance of efficiency and accuracy for a complex, multi-compartment, multi-dataset, and multi-metabolite PBPK model. This method was first introduced in @cukier1973study (the classic FAST method with first-order effect only) in a study of the sensitivity of coupled reaction systems that are constructed by differential equations with the numerous rate coefficients. It then evolved to Sobol' sensitivity measure in @saltelli1999quantitative (the eFAST method with first- and total-order effects). The FAST algorithm creates the multidimensional space of input parameters through a "search-curve". It's more computationally efficient in calculating the influence of model parameter than other global SA methods such as the Monte Carlo based approaches [@jansen1999analysis; @owen2013better]. Our previous work also found some efficient visualization approaches that can be used to distinguish between “influential” and “non-influential” parameters. This approach addresses common identifiability issues of PBPK model parameter that increase the computational burden of Bayesian analysis [@garcia2015identifiability], and therefore reduce the computational efficiency. We also developed approaches for communicating parameter sensitivity in decision making.

@@

Although these tools provide various approaches to perform SA, they each individually have limitations for application to pharmacokinetic and PK modeling. For example, the estimated index in SA is not robust when using the inadequate size of the sample number. However, there are no software packages that provide the available tools in the assessment of reliability. In addition, the existing eFAST function does not includes the feature to generate the random sampling curve. We present here an \proglang{R} package, called \pkg{pksensi}, which is designed to make SA more accessible and reproducible in pharmacological and toxicological researches. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic and PK model with multivariate model outputs. Moreover, it seamlessly integrates with both general ODE solvers available in R or as external \proglang{C} code, as well as the \pkg{GNU MCSim} software used for conducting sophisticated PK modeling, including MCMC. The analytical framework in \pkg{pksensi} not only includes global SA but also the uncertainty analysis and diagnostic tools to support the modeling process. The design concepts of \pkg{pksensi} are:

*	Cross-platform: Models can run on Windows/MacOS/Linux
*	Freely available: All related packages are free and open source
*	Integrated application: Users can run pharmacokinetic models in \proglang{R} with script that were written in \proglang{C} or \pkg{GNU MCSim} 
*	Decision support: The output results and visualization tools can be used to easily determine which parameters have "non-influential" effects on the model output and can be fixed in model calibration.

@@

We adopted eFAST, a widely used analysis of variance (ANOVA)-like global SA approach to decompose the variance to partial variances for multidimensional parameters in numerical models in this package. Most of the available eFAST functions, such as the algorithm to generate the parameter space and to set the sampling frequency, were sourced from the \proglang{R} \pkg{sensitivity} package [@R-sensitivity]. The eFAST method computes the sensitivity index (SI) via ANOVA-like decomposition of the function for analysis. This type of global SA approach and SI also know as Sobol' method and Sobol' index. Since the estimated SI is not stable (high variation) when using smaller sample sizes, we included a random phase-shift approach to conduct replicating sampling from random starting points across parameter space to test the convergence and robustness of the sensitivity measurement. Considering the model that is built under $y=f(x_{i})$. The sampling scheme can be written as,

@@

where $x_i$ is the nominal value of the $i$-th parameter, and $\omega_i$ is a vector giving the set of frequencies, one frequency for each parameter, and $\varphi$ is a random phase-shift coefficient ranged from 0 to $2\pi$. The default set of frequencies is based on the study from @saltelli1999quantitative. Through the random phase-shift, the robust result of the sensitivity measurement should be similar across replications under the same sample size.
For most Sobol'-based SA algorithm, the larger sample size is needed to reduce the probability of having negative estimates. However, FAST estimates are always positive according to its construction.

To investigate the convergence of SI, we adopted the "exam" approach proposed by [@sarrazin2016global]. This method quantitatively assesses convergence by computing the width of 95% confidence intervals of computed SI for all parameters across all time-points and output variables. Through random phase shift procedure, the replicate SI values can be repeatedly sampled independently. Therefore, they can be used to investigate the convergence of SI as,

@@

The R package \pkg{pksensi} provides a useful function to solve ODEs in pharmacokinetic and PK models using analytical and numerical approaches. The solution can be performed under the pure \proglang{R} programming environment by linking pksensi with the \pkg{deSolve} package [@JSSv033i09]. However, the most efficient way to achieve high computational speed is through compiled, lower-level languages, such as \proglang{FORTRAN}, \proglang{C}, or \proglang{C++}. The \pkg{pksensi} includes a function to compile and create dynamic-link libraries (.dll) on Windows and shared objects (.so) on Unix-liked systems (e.g., Linux and MacOS). This compiled program can load and execute through the build-in function in \proglang{R}. However, Windows users need to install Rtools to compile the source code by using the GNU GCC compiler. More requirements are detailed in @soetaertr. The \proglang{C} implementation of the example model can be found as an example within \pkg{GNU MCSim} [@JSSv002i09]. The \pkg{pksensi} can also link with \pkg{GNU MCSim} to create the model program, used in solving each system of equations.

The uncertainty analysis is a crucial modeling process within SA. Sometimes, it's difficult to have informative data to set up the parameter distribution. A wider range for a parameter distribution might cause a numerical error at extreme values. On the other hand, narrower ranges might cause the incorrect calibrated result of the model parameter. The uncertainty analysis using Monte Carlo is an appropriate approach in the data-driven modeling process to address this problem. Before fitting the data to model, the Monte Carlo can be applied with the given distribution of model parameter to examine whether the experiment data within the range of in the distribution of model output. \pkg{pksensi} also has a function that can link to \pkg{GNU MCSim} to conduct Monte Carlo analysis which is built in \pkg{GNU MCSim} and exam the simulation result. The general workflow of uncertainty and sensitivity analysis is illustrated in Figure \ref{fig:Workflow}.

@@

The output of SA returns a list that contains the given time-points and values of all state variables (e.g., chemical concentrations in blood). This format is particularly suited for further graphical routines within \pkg{pksensi}. Also, time-dependent sensitivity measurements of first and total order effects for all parameters can be plotted simultaneously. If the computed SI show different values across replications under the same sample size, this indicates inadequate sample number, which can create an unreliable results. It is essential to use a sufficient sample size to prevent incorrect judgments in parameter fixing. \pkg{pksensi} includes functions to check the convergence and sensitivity of model parameters, providing a means to assess the robustness of the sensitivity measurement. We also developed a "cut-off"-based approach in this package to distinguish between "influential" and "non-influential" parameters. Finally, \pkg{pksensi} provides visualization tools for the effective investigation and communication of results in decision support.

@@

In this section, two PK models were applied to demonstrate how \pkg{pksensi} works. The first one is a generic model and the second is a chemical-specific model. The following code shows the version of the platform, version of \proglang{R} and \pkg{pksensi} that use to conduct the current example. We use \pkg{GNU MCSim} v6.1.0 in these examples.

@@

In the first example, we use a generic, one-compartment PK model from \pkg{httk} package [@JSSv079i04] to demonstrate how \pkg{pksensi} can be applied to PK studies. The differential equations for this model is as,

@@

The concentration of the chemical in the rest of body and blood ($C_{rest}$, mg/L) can be calculated as,

@@

where $V_{dist}$ is the volume of distribution (L/kg BW) and $BW$ is the body weight (kg). The $C_{rest}$ can also be seen as the chemical concentration in plasma that can be further used to compare with observed results in a pharmacokinetic experiment. The bioavailability is assumed to be 100% in this model.

@@

To start, we implemented the one-compartment PK model in \proglang{R}. \pkg{pksensi} allows users select the preferred method to solve the PK model, either with the \pkg{deSolve} package or with \pkg{GNU MCSim} through the compile function. This section mainly focus on how to conduct global SA with \pkg{deSolve} package with pure \proglang{R} environment.

@@

The parameter values and initial states need to assign to the specific value before simulation. We used the corresponding parameter value of acetaminophen (APAP) in this example. These model parameters are derived from the *in-vivo* or *in-vitro* experiment results. The parameter value can be generated from \code{parameterize_1comp} function in \pkg{httk} package as:

@@

We conduct global SA for the four parameters in one-compartment PK model to investigate the parameter impact on the plasma concentration during the 24-hr time period post intake. The distribution of parameter is taken to be uniform with bounds corresponding to 50% and 200% of the nominal value. Therefore, the parameter ranges are assumed to be 0.55 and 2.2 L/kg BW for \code{vdist}. The \code{ke} are ranged from 0.12 to 0.46 /h, corresponding to half-times of 1.5 and 5.8 hr. The \code{ka} are ranged 1.09 to 4.36 /h, corresponding to half-times of 0.32 and 0.64 hr. The \code{BW} is assumed to a normal distribution with mean = 60 kg and sd = 5 kg.

@@

The generated parameters are stored as a 3-D array under the named \code{a}, with the dimension of sample size, number of replications, and number of parameters, respectively.

@@

The sample number is 200 with 4 model parameters, which generates 800 model evaluations. The replication is set to 10. Therefore, the total of 8,000 parameter sequence will be used to compute the corresponding outputs. Figure \ref{fig:rep-sampl} plotted the sampling process for each parameter from the first 3 replications. The search curves show the different intensity of sampling patterns in each segment.

@@

Here, we can see (Figure \ref{fig:pk-solvefun-rest}) that \code{vdist} and \code{ke} are dominating the plasma concentration before and after 5-hr post chemical intake, respectively. Besides, the \code{kgutabs} only plays a crucial role to determine the plasma concentration in the first hour. However, the current result is only based on the setting distribution of model parameters for APAP. The different given input conditions (e.g., range of parameter uncertainty, chemical-dependent parameter value) can change the result (result not shown).

@@

Figure \ref{fig:pk-solvefun-Amet} shows that the amount of metabolized is also determined by parameter \code{ke}. Same as \code{Ccompartment}, the \code{kgutabs} contribute about 30 - 40% variation of model output in the first hour. The \code{BW} is the least important parameter in the current analysis, and therefore can be fixed in the model fitting to data and additional applications.

In addition to use time-SI profile to investigate the parameter impact on model output, we can construct the relationship between parameter and the corresponding value of model output in the examination.

@@

The output variable \code{out} containing all the input arguments detailed before and the calculated SI of first order (\code{mSI}), interaction (\code{iSI}), and total order (\code{tSI}). The convergence indices are also stored in the list named \code{mCI}, \code{iCI}, and \code{tCI}. The output are formatted as 4-D array in \code{y} with the dimension name of model evaluation, number of replications, number of time points, and number of output variables, respectively.

@@

Some functions in \pkg{pksensi} provide an efficient way to check the result from global SA. The \code{check} can determine which parameters have relatively lower sensitivity measurement across the given time points and model outputs, and therefore can be applied parameter fixing in model calibration. The \code{check} also provides some feasible argument to specify the target output or change the cut-off value. The argument of \code{SI.cutoff} is setting at 0.05 to detect the relative non-influential parameters as default, representing the 5% change of the output is contributed from the specific parameter variation.

@@

The computing time of using \code{solve_fun} in SA is estimated as,

@@

Then, before we conduct the SA through \pkg{GNU MCSim}, The following code is used to compile the \pkg{GNU MCSim} model code to the executable program. To solve ODEs through \pkg{GNU MCSim}, we need to use \code{compile_model} with argument \code{application = mcsim}.

@@

Similar to \code{solve_fun} function that can define the initial value of parameters and state variable through generated functions, the \code{solve_mcsim} also has a \code{condition} argument that can be used to give the specific input value such as exposure dose, fixed parameter value or initial condition of state variables.

@@

The aim of this section is to apply global SA for APAP-PBPK model through \pkg{pksensi}. The model describes the pharmacokinetics of parent APAP and its two metabolites glucuronide (APAP-G) and sulfate (APAP-S). The model code is included in this package in https://github.com/nanhung/pksensi/blob/master/tests/pbpk_apap.model and can be generated through \code{pbpk_apap_model} function. We applied the global SA workflow to the original published model with 21 model parameters [@s13318-015-0253-x]. The descriptions of each parameter and the sampling ranges are list in Table \ref{tab:data}.

@@

As in the example of the one-compartment PK model, the model parameter and the corresponding sampling range should be defined to create the parameter matrix. Previously, the probability distributions of model parameters were set to either truncated normal or uniform distributions when the parameters have informative prior information or not, respectively. To rapidly reach the acceptance convergence in this example, we apply uniform distribution for all testing parameters. The ranges of informative parameters are set to 1.96-times difference for the single side under log-scaled (approximate 54.6 times difference between minimum and maximum in natural scaled). The nominal values of informative model parameters were defined as:

@@

Generally, wide range of parameter value might cause the computing error when solving the differential equation. One of the effective ways to prevent this problem is to adjust the value of relative and absolute error tolerance to control the error appearance by resetting these parameters in a lower value. The \code{generate_infile} and \code{solve_mcsim} provide the arguments of \code{rtol} and \code{atol} that adjust the error tolerance to prevent the unwanted error. However, the modification will decrease the computing speed. Therefore, the alternative method to prevent this issue is to detect the crucial parameter range that causes the problem. Also, setting the maximum number of steps to higher value in \pkg{GNU MCSim} can prevent this problem (internally defined). The maximum number of step is set to 5000 (default) in this case. The value can reset through reinstall the the \pkg{GNU MCSim}. Here we separate the global SA of APAP-PBPK model process to several steps.

@@

The model code needs to be prepared in the following global SA workflow. After creating the \code{pbpk_apap.model} file in the working directory, the next step is to generate the executable program (\code{mcsim.pbpk_apap}) through the \code{compile_model} function.

@@

To optimize the computing speed, this case only uses \pkg{GNU MCSim} to estimate the concentration of APAP and its metabolites in plasma. The oral dose of APAP is 20 mg/kg in this example. Generally, the input dosing method can be defined through the \code{condition} argument. Since the unit of the given dose is mg/kg, the \code{mgkg_flag} is set to 1. Additional input schedule functions can be found in the section of input functions in \pkg{GNU MCSim} User’s Manual (https://www.gnu.org/software/mcsim/mcsim.html#Input-functions).

@@

Based on our previous study, we proposed the heatmap visualization approach \code{heat_check} to distinguish "influential" and "non-influential" parameters with a "cut-off" point. Through the given argument \code{order}, we can select the specific order of sensitivity measurement that we're interested in (Figure \ref{fig:Heat-SI}).

@@

Adding the `index = "CI"` in the function can further investigate the convergence index. Based on the current setting of sampling size, most parameters cannot reach the acceptable criteria of convergence (Figure \ref{fig:Heat-CI}). Therefore, a higher number of sampling is necessary. The sample size of convergence in the current PBPK model is 8,192 [@fphar201800588]. However, based on the current sample size we still can find 6 parameters that can be an important parameter for the plasma APAP concentration.

@@

Computational modeling has become an important technique in modern scientific research to support hypotheses, experiments, and discoveries.  Uncertainty and SA are essential tools for modelers in the conduct of model development and examination. Several packages on the CRAN repository and other software platforms had been built to perform global SA with a variety of algorithms, but none have included integration with ODE-based models, such as PK models.

We initiated \pkg{pksensi} in order to facilitate a comprehensive and efficient global SA workflow in PK modeling, filling a key gap for open source modeling communities. It provides a straightforward application to investigate the impact of parameters on model outputs across simulation time-points and variables. We adopted the eFAST method, a common variance-based SA that we found to have the best balance of efficiency and accuracy. In addition, the package also includes the ability to assess the convergence of SA results, which is rarely addressed in most global SA studies and software packages. We also developed functions to visualize the output results and help distinguish the "influential" and "non-influential" parameters that can be applied in parameter fixing in the model calibration. In the future version, we expect to add additional global SA approaches [@glen2012estimating; @ring2017identifying]. The other features such as principle component analysis [@lamboni2011multivariate] will also be considered.

@@

It is worth noting that the current global SA results are conditional on the specific model and the given range of the model parameters. Other factors, such as the number of parameters examined, the parameter distributions/ranges, and the given dosing level in PK model might also have a potential impact on the level of importance of each parameter, but these "meta-uncertainties" are not currently addressed. The complex models with a high dimensionality in the the parameter is also a challenge in global SA as they have higher computer demands. As approaches are developed to address these complexities, we hope to incorporate them into \pkg{pksensi}.

Overall, \pkg{pksensi} provides a comprehensive suite features and functions for performing global SA for PK models and can create robust and reproducible results for decision making in model development and calibration. Although \pkg{pksensi} used here is mainly for pharmacokinetic modeling, it can also be applied to other ODE-based dynamic models in order to investigate the sensitivity of model outputs to input parameters.

>>>>

Sensitivity analysis (SA) is a mathematical technique to investigate how variations in model parameters affect model outputs [@saltelli2008global]. An increasing number of studies use SA approach to determine which model parameters contribute to high variation in model predictions [@ferretti2016trends]. This technique has also been applied in pharmacology and toxicology research [@loizou2015application; @mcnally2012reconstruction]. Pharmacokinetic models describe the changes in the concentrations or amounts of a substance in the body over time (the various terms for the kinetic models, including "pharmacokinetic", "toxicokinetic", and "biokinetic" models, are used exchangeably here). The goal of SA in pharmacokinetic research is to examine the sensitivity of output variables, such as chemical concentration in blood or tissues, with respect to input parameters, such as anatomical, physiological, and kinetic constants [@mcnally2011workflow]. In addition, this SA approach can guide experimental design and the parameter estimation processes [@zhang2015sobol]. For instance, SA can identify "unidentifiable" parameters that can lead to problems with numerical procedures used in parameter estimation [@chu2010quantitative]. It can be further applied to parameter prioritization and parameter fixing before model calibration [@fphar201800588].

In our earlier work [@fphar201800588], we developed an updated workflow to apply global SA to reduce the computational burden in the Bayesian, Markov Chain Monte Carlo (MCMC)-based calibration process of a physiologically based pharmacokinetic (PBPK) model. We used \pkg{GNU MCSim} [@JSSv002i09], an effective simulation package for Bayesian population PBPK modeling, to calibrate the model. We found that the extended Fourier Amplitude Sensitivity Test (eFAST), a type of variance-based global SA algorithm, had the best balance of efficiency and accuracy for a sophisticated, multi-compartment, multi-dataset, and multi-metabolite PBPK model. This method was first introduced in @cukier1973study (the classic FAST method with the first-order effect only) in a study of the sensitivity of coupled reaction systems that are constructed by differential equations with the numerous rate coefficients. It then evolved to Sobol' sensitivity measure in @saltelli1999quantitative (the eFAST method with first- and total-order effects). The FAST algorithm creates the multidimensional space of input parameters through a "search-curve". It is more computationally efficient in calculating the influence of model parameter than other global SA methods such as the Monte Carlo based approaches [@jansen1999analysis; @owen2013better]. Our previous work also found some efficient visualization approaches that can be used to distinguish between “influential” and “non-influential” parameters. This approach addresses common identifiability issues of PBPK model parameter that increase the computational burden of Bayesian analysis [@garcia2015identifiability], and therefore reduce the computational efficiency. We also developed approaches for communicating parameter sensitivity in decision making.

@@

Although these tools provide various approaches to perform SA, they each individually have limitations for application to pharmacokinetic and PK modeling. For example, the estimated index in SA is not robust when using the inadequate size of the sample number. However, there are no software packages that provide the available tools in the assessment of reliability. In addition, the existing eFAST function does not include the feature to generate the random sampling curve. We present here an \proglang{R} package, called \pkg{pksensi}, which is designed to make SA more accessible and reproducible in pharmacological and toxicological researches. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic and PK model with multivariate model outputs. Moreover, it seamlessly integrates with both general ODE solvers available in R or as external \proglang{C} code, as well as the \pkg{GNU MCSim} software used for conducting sophisticated PK modeling, including MCMC. The analytical framework in \pkg{pksensi} includes not only global SA but also the uncertainty analysis and diagnostic tools to support the modeling process. The design concepts of \pkg{pksensi} are:

*	Cross-platform: Models can run on Windows/MacOS/Linux
*	Freely available: All related packages are free and open-source
*	Integrated application: Users can run pharmacokinetic models in \proglang{R} with script that were written in \proglang{C} or \pkg{GNU MCSim} 
*	Decision support: The output results and visualization tools can be used to easily determine which parameters have "non-influential" effects on the model output and can be fixed in model calibration.

@@

We adopted eFAST, the widely used analysis of variance (ANOVA)-like global SA approach to decompose the variance to partial variances for multidimensional parameters in numerical models in this package. Most of the available eFAST functions, such as the algorithm to generate the parameter space and to set the sampling frequency, were sourced from the \proglang{R} \pkg{sensitivity} package [@R-sensitivity]. The eFAST method computes the sensitivity index (SI) via ANOVA-like decomposition of the function for analysis. This type of global SA approach and SI also know as Sobol' method and Sobol' index. Since the estimated SI is not stable (high variation) when using smaller sample sizes, we included a random phase-shift approach to conduct replicating sampling from random starting points across parameter space to test the convergence and robustness of the sensitivity measurement. Considering the model that is built under $y=f(x_{i})$. The sampling scheme can be written as,

@@

where $x_i$ is the nominal value of the $i$-th parameter, and $\omega_i$ is a vector giving the set of frequencies, one frequency for each parameter, and $\varphi$ is a random phase-shift coefficient ranged from 0 to $2\pi$. The default set of frequencies is according to the study from @saltelli1999quantitative. Through the random phase-shift, the robust result of the sensitivity measurement should be similar across replications under the same sample size.
For most Sobol'-based SA algorithm, the larger sample size is needed to reduce the probability of having negative estimates. However, FAST estimates are always positive according to its construction.

To investigate the convergence of SI, we adopted the "exam" approach proposed by [@sarrazin2016global]. This method quantitatively assesses convergence by computing the width of 95% confidence intervals of computed SI for all parameters across all time-points and output variables. Through random phase shift procedure, the replicated SI values can repeatedly sample independently. Therefore, they can be used to investigate the convergence of SI as,

@@

The R package \pkg{pksensi} provides a useful function to solve ODEs in pharmacokinetic and PK models using analytical and numerical approaches. The solution can perform under the pure \proglang{R} programming environment by linking pksensi with the \pkg{deSolve} package [@JSSv033i09]. However, the most efficient way to achieve high computational speed is through compiled, lower-level languages, such as \proglang{FORTRAN}, \proglang{C}, or \proglang{C++}. The \pkg{pksensi} includes a function to compile and create dynamic-link libraries (.dll) on Windows and shared objects (.so) on Unix-liked systems (e.g., Linux and MacOS). This compiled program can load and execute through the build-in function in \proglang{R}. However, Windows users need to install Rtools to compile the source code by using the GNU GCC compiler. More requirements were detailed in @soetaertr. The \proglang{C} implementation of the example model can be found as an example within \pkg{GNU MCSim} [@JSSv002i09]. The \pkg{pksensi} can also link with \pkg{GNU MCSim} to create the model program, used in solving each system of equations.

The uncertainty analysis is a crucial modeling process within SA. Sometimes, it is difficult to have informative data to set up the parameter distribution. A wider range for a parameter distribution might cause a numerical error at extreme values. On the other hand, narrower ranges might cause the incorrect calibrated result of the model parameter. The uncertainty analysis using Monte Carlo is an appropriate approach in the data-driven modeling process to address this problem. Before fitting the data to model, the Monte Carlo can be applied with the given distribution of model parameter to examine whether the experiment data within the range of in the distribution of model output. \pkg{pksensi} also has a function that can link to \pkg{GNU MCSim} to conduct Monte Carlo analysis which is built-in \pkg{GNU MCSim} and exam the simulation result. The general workflow of uncertainty and SA is illustrated in Figure \ref{fig:Workflow}.

@@

The output of SA returns a list that contains the given time-points and values of all state variables (e.g., chemical concentrations in blood). This format is particularly suited for further graphical routines within \pkg{pksensi}. Also, time-dependent sensitivity measurements of first and total effects for all parameters can be plotted simultaneously. If the computed SI show different values across replications under the same sample size, this indicates an inadequate sample number, which can create an unreliable result. It is essential to use a sufficient sample size to prevent incorrect judgments in parameter fixing. \pkg{pksensi} includes functions to check the convergence and sensitivity of model parameters, providing a means to assess the robustness of the sensitivity measurement. We also developed a "cut-off"-based approach in this package to distinguish between "influential" and "non-influential" parameters. Finally, \pkg{pksensi} provides visualization tools for the effective investigation and communication of results in decision support.

@@

In this section, two PK models will apply to demonstrate how \pkg{pksensi} works. The first one is a generic model, and the second is a chemical-specific model. The following code shows the version of the platform, version of \proglang{R}, and \pkg{pksensi} that use to conduct the current example. We use \pkg{GNU MCSim} v6.1.0 in these examples.

@@

In the first example, we use a generic, one-compartment PK model from \pkg{httk} package [@JSSv079i04] to demonstrate how \pkg{pksensi} can apply to PK studies. The differential equations for this model is as,

@@

The concentration of the chemical in the rest of body and blood ($C_{rest}$, mg/L) can calculate as,

@@

where $V_{dist}$ is the volume of distribution (L/kg BW) and $BW$ is the body weight (kg). The $C_{rest}$ can also represent as the chemical concentration in plasma that can be further used to compare with observed results in a pharmacokinetic experiment. The bioavailability is assumed to be 100% in this model.

@@

To start, we implemented the one-compartment PK model in \proglang{R}. \pkg{pksensi} allows users select the preferred method to solve the PK model, either with the \pkg{deSolve} package or with \pkg{GNU MCSim} through the compile function. This section mainly focuses on how to conduct global SA with \pkg{deSolve} package with pure \proglang{R} environment.

@@

The parameter values and initial states need to assign to the specific value before simulation. Here, we use the corresponding parameter value of acetaminophen (APAP) in this example. These model parameters are derived from the *in-vivo* or *in-vitro* experiment results. The parameter value can be generated from \code{parameterize_1comp} function in \pkg{httk} package as:

@@

We conduct global SA for the four parameters in one-compartment PK model to investigate the parameter impact on the plasma concentration during the 24-hr time post intake. The distribution of parameter is taken to be uniform with bounds corresponding to 50% and 200% of the nominal value. Therefore, the parameter ranges are assumed to be 0.55 and 2.2 L/kg BW for \code{vdist}. The \code{ke} are ranged from 0.12 to 0.46 /h, corresponding to half-times of 1.5 and 5.8 hr. The \code{ka} are ranged 1.09 to 4.36 /h, corresponding to half-times of 0.32 and 0.64 hr. The \code{BW} is assumed to a normal distribution with mean = 60 kg and sd = 5 kg.

@@

The generated parameters are stored as a 3-D array under the named \code{a}, with the dimension of sample size, the number of replications, and the number of parameters, respectively.

@@

The sample number is 200, with 4 model parameters, which generates 800 model evaluations. The replication is set to 10. Therefore, the total of 8,000 parameter sequence will be used to compute the corresponding outputs. Figure \ref{fig:rep-sampl} plotted the sampling process for each parameter from the first 3 replications. The search curves show the different intensity of sampling patterns in each segment.

@@

Here, we can see (Figure \ref{fig:pk-solvefun-rest}) that \code{vdist} and \code{ke} dominate the plasma concentration before and after 5-hr post chemical intake, respectively. Besides, the \code{kgutabs} only plays a crucial role to determine the plasma concentration in the first hour. However, the current result only based on the setting distribution of model parameters for APAP. The different given input conditions (e.g., range of parameter uncertainty, chemical-dependent parameter value) can change the result (result not shown).

@@

Figure \ref{fig:pk-solvefun-Amet} shows that the amount of metabolized is also determined by parameter \code{ke}. Same as \code{Ccompartment}, the \code{kgutabs} contribute about 30 - 40% variation of model output in the first hour. The \code{BW} is the least important parameter in the current analysis, and therefore, can be fixed in the model fitting to data and additional applications.

Also, to use time-SI profile to investigate the parameter impact on model output, we can construct the relationship between parameter and the corresponding value of model output in the examination.

@@

The output variable \code{out} containing all the input arguments detailed before and the calculated SI of first order (\code{mSI}), interaction (\code{iSI}), and total order (\code{tSI}). Convergence indices are also stored in the list named \code{mCI}, \code{iCI}, and \code{tCI}. The outputs are formatted as 4-D array in \code{y} with the dimension name of model evaluation, number of replications, number of time points, and number of output variables, respectively.

@@

Some functions in \pkg{pksensi} provide efficient ways to check the result from global SA. The \code{check} can determine which parameters have relatively lowered sensitivity measurement across the given time points and model outputs, and therefore can be applied parameter fixing in model calibration. The \code{check} also provides some feasible argument to specify the target output or change the cut-off value. The argument of \code{SI.cutoff} is setting at 0.05 to detect the relative non-influential parameters as default, representing the 5% change of the output is contributed from the specific parameter variation.

@@

The computing time of using \code{solve_fun} in SA estimate as,

@@

Then, before we conduct the SA through \pkg{GNU MCSim}, The following codes are used to compile the \pkg{GNU MCSim} model code to the executable program. To solve ODEs through \pkg{GNU MCSim}, we need to use \code{compile_model} with argument \code{application = mcsim}.

@@

Similar to \code{solve_fun} function that can define the initial values of parameters and state variable through generated functions, the \code{solve_mcsim} also has a \code{condition} argument that can be used to give the specific input value such as exposure dose, nominated parameter value, or initial condition of state variables.

@@

This section aims to apply global SA for APAP-PBPK model through \pkg{pksensi}. The model describes the pharmacokinetics of parent APAP and its two metabolites glucuronide (APAP-G) and sulfate (APAP-S). The model code is included in this package in https://github.com/nanhung/pksensi/blob/master/tests/pbpk_apap.model and can generate through \code{pbpk_apap_model} function. We apply the global SA workflow to the original published model with 21 model parameters [@s13318-015-0253-x]. The descriptions of each parameter and the sampling range are list in Table \ref{tab:data}.

@@

As in the example of the one-compartment PK model, the model parameter and the corresponding sampling range should be defined to create the parameter matrix. Previously, the probability distributions of model parameters were set to either truncated normal or uniform distributions when the parameters have informative prior information or not, respectively. To rapidly reach the acceptance convergence in this example, we apply uniform distribution for all testing parameters. The ranges of informative parameters set to 1.96-times difference for the single side under log-scaled (approximate 54.6 times difference between minimum and maximum in natural scaled). The nominal values of informative model parameters define as:

@@

Generally, a wide range of parameter value might cause the computing error when solving the differential equation. One of the effective ways to prevent this problem is to adjust the value of relative and absolute error tolerance to control the error appearance by resetting these parameters in a lower value. The \code{generate_infile} and \code{solve_mcsim} provide the arguments of \code{rtol} and \code{atol} that adjust the error tolerance to prevent the unwanted error. However, the modification will decrease the computing speed. Therefore, the alternative method to prevent this issue is to detect the crucial parameter range that causes the problem. Also, setting the maximum number of steps to the higher value in \pkg{GNU MCSim} can prevent this problem (internally defined). The maximum number of step is 5000 (default) in this case. The value can reset through reinstall the \pkg{GNU MCSim}. Here we separate the global SA of APAP-PBPK model process to several steps.

@@

The model code needs to prepare in the following global SA workflow. After creating the \code{pbpk_apap.model} file in the working directory, the next step is to generate the executable program (\code{mcsim.pbpk_apap}) through the \code{compile_model} function.

@@

In this case, we only use \pkg{GNU MCSim} to estimate the concentration of APAP and its metabolites in plasma to optimize the computing speed. The oral dose of APAP is 20 mg/kg in this example. Generally, the input dosing method can be defined through the \code{condition} argument. Since the unit of the given dose is mg/kg, the \code{mgkg_flag} is set to 1. Additional input schedule functions can be found in the section of input functions in \pkg{GNU MCSim} User’s Manual (https://www.gnu.org/software/mcsim/mcsim.html#Input-functions).

@@

Based on our previous study, we proposed the heatmap visualization approach \code{heat_check} to distinguish "influential" and "non-influential" parameters with a "cut-off" point. Through the given argument \code{order}, we can select the specific order of sensitivity measurement that we are interested in (Figure \ref{fig:Heat-SI}).

@@

Adding the `index = "CI"` in the function can further investigate the convergence index. Based on the current setting of sampling size, most parameters cannot reach the acceptable criteria of convergence (Figure \ref{fig:Heat-CI}). Therefore, a higher number of sampling is necessary. The sample size of convergence in the current PBPK model is 8,192 [@fphar201800588]. However, based on the current sample size, we still can find 6 parameters that can be an important parameter for the plasma APAP concentration.

@@

Computational modeling has become an essential technique in modern scientific research to support hypotheses, experiments, and discoveries. Uncertainty and SA are essential tools for modelers in the conduct of model development and examination. Several packages on the CRAN repository and other software platforms had been built to perform global SA with a variety of algorithms, but none have included integration with ODE-based models, such as PK models.

We initiated \pkg{pksensi} in order to facilitate a comprehensive and efficient global SA workflow in PK modeling, filling a crucial gap for open-source modeling communities. It provides a straightforward application to investigate the impact of parameters on model outputs across simulation time-points and variables. We adopted the eFAST method, a common variance-based SA that we found to have the best balance of efficiency and accuracy. In addition, the package also includes the ability to assess the convergence of SA results, which is rarely addressed in most global SA studies and software packages. We also developed functions to visualize the output results and help distinguish the "influential" and "non-influential" parameters that can be applied in parameter fixing in the model calibration. In the future version, we expect to add additional global SA approaches [@glen2012estimating; @ring2017identifying]. The other features, such as principle component analysis [@lamboni2011multivariate] will also be considered.

@@

It is worth noting that the current global SA results are conditional on the specific model and the given range of the model parameters. Other factors, such as the number of parameters examined, the parameter distributions/ranges, and the given dosing level in PK model might also have a potential impact on the level of importance of each parameter, but these "meta-uncertainties" are not currently addressed. The complex models with high dimensionality in the parameter is also a challenge in global SA as they have higher computer demands. As approaches are developed to address these complexities, we hope to incorporate them into \pkg{pksensi}.

Overall, \pkg{pksensi} provides full suite features and functions for performing global SA for PK models and can create robust and reproducible results for decision making in model development and calibration. Although \pkg{pksensi} used here is mainly for pharmacokinetic modeling, it can also be applied to other ODE-based dynamic models in order to investigate the sensitivity of model outputs to input parameters.
